---
title: "Local Authority Trend Analysis"
output: github_document
---

This short analysis is to highlight the variation in crashes affecting active traveling across Great Britain, and to try to understand the variation in risk between different Local Authorities.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(sf)
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyr)
library(tmap)
```

## Getting Data

First stage is to get the crash data, in this analysis we used the stats19 data from 2010 to 2019 (the most recent 10 years) which has been adjusted to account for the different causality reporting systems used over that time.

We also use the 2019 Local authority boundaries and the 2019 population estimates, for the workplace population we must used the 2011 census.


```{r data_setup, include=FALSE}
if(!dir.exists("data")){
  dir.create("data")
}

if(!file.exists("data/casualties_adjusted.Rds")){
  download.file(piggyback::pb_download_url("casualties_adjusted.Rds", tag = "v0.1"),
                "data/casualties_adjusted.Rds", mode = "wb")
}

if(!file.exists("data/crashes_2010_2019.Rds")){
  download.file(piggyback::pb_download_url("ac10.Rds", tag = "0.1"),
                "data/crashes_2010_2019.Rds", mode = "wb")
}

if(!file.exists("data/counties_uas_gb_2019_ubc.Rds")){
  download.file(piggyback::pb_download_url("counties_uas_gb_2019_ubc.Rds", tag = "0.1.1"),
              destfile = "data/counties_uas_gb_2019_ubc.Rds", mode = "wb")
}

if(!file.exists("data/upper_LA_population_2019.csv")){
  download.file(piggyback::pb_download_url("upper_LA_population_2019.csv", tag = "0.1.1"),
              destfile = "data/upper_LA_population_2019.csv", mode = "wb")
}

if(!file.exists("data/upper_LA_work_population.csv")){
  download.file(piggyback::pb_download_url("upper_LA_work_population.csv", tag = "0.1.1"),
              destfile = "data/upper_LA_work_population.csv", mode = "wb")
}

# File doesn't exist - no valid source for Scotland's workplace population
# if(!file.exists("data/scot_work_population.csv")){
#   download.file(piggyback::pb_download_url("scot_work_population.csv", tag = "0.1.1"),
#               destfile = "data/scot_work_population.csv", mode = "wb")
# }

if(!file.exists("data/bdline_gpkg_gb.zip")){
  download.file("https://api.os.uk/downloads/v1/products/BoundaryLine/downloads?area=GB&format=GeoPackage&redirect",
                destfile = "data/bdline_gpkg_gb.zip", mode = "wb")
}



acc <- readRDS("data/crashes_2010_2019.Rds")

cas <- readRDS("data/casualties_adjusted.Rds")

la_raw <- readRDS("data/counties_uas_gb_2019_ubc.Rds")
la_raw$ctyua19cd[la_raw$ctyua19cd == "E10000002"] <- "E06000060"
la_raw <- la_raw[,c("ctyua19cd","ctyua19nm")]
names(la_raw) <- c("code","la_name","geometry")

pop <- read.csv("data/upper_LA_population_2019.csv")
pop <- pop[pop$mnemonic %in% la_raw$code,]
pop <- pop[,2:3]
names(pop) <- c("code","population_2019")

pop_work_ew <- read.csv("data/upper_LA_work_population.csv")
pop_work_ew <- pop_work_ew[,c(3,4)]
names(pop_work_ew) <- c("code","work_pop_2011")

# file doesn't exist
# pop_work_sct <- read.csv("data/scot_work_population.csv")
# pop_work_sct <- pop_work_sct[6:358,]
# names(pop_work_sct) <- c("id","work_pop")
# pop_work_sct$code <- substr(pop_work_sct$id,1,9)

# Fix old codes
pop_work_ew$code[pop_work_ew$code == "E10000002"] <- "E06000060"
pop_work_ew$code[pop_work_ew$code == "E06000028"] <- "E06000058"
pop_work_ew$code[pop_work_ew$code == "E06000029"] <- "E06000058"
pop_work_ew$code[pop_work_ew$code == "E10000009"] <- "E06000059"

pop_work_ew <- pop_work_ew %>% 
  group_by(code) %>%
  summarise(work_pop_2011 = sum(work_pop_2011))

# dir.create("tmp")
# unzip("data/bdline_gpkg_gb.zip", exdir = "tmp")
# bounds_scot <- read_sf("tmp/data/bdline_gb.gpkg")

```


```{r get_data, echo = FALSE, include=FALSE, warning=FALSE}
# Join ACC and CAS and summarize
acc <- acc[!is.na(acc$longitude),]
acc <- st_as_sf(acc, coords = c("longitude","latitude"), crs = 4326)
acc$location_easting_osgr <- NULL
acc$location_northing_osgr <- NULL

cas_summary <- cas %>%
  group_by(accident_index) %>%
  summarise(casualty_slight = sum(Adjusted_Slight, na.rm = TRUE),
            casualty_serious = sum(Adjusted_Serious, na.rm = TRUE),
            casualty_fatal = length(casualty_severity[casualty_severity == "Fatal"]),
            casualty_slight_pedestrian = sum(Adjusted_Slight[casualty_type == "Pedestrian"], na.rm = TRUE),
            casualty_serious_pedestrian = sum(Adjusted_Serious[casualty_type == "Pedestrian"], na.rm = TRUE),
            casualty_fatal_pedestrian = length(casualty_severity[casualty_severity == "Fatal" & casualty_type == "Pedestrian"]),
            casualty_slight_cyclist = sum(Adjusted_Slight[casualty_type == "Cyclist"], na.rm = TRUE),
            casualty_serious_cyclist = sum(Adjusted_Serious[casualty_type == "Cyclist"], na.rm = TRUE),
            casualty_fatal_cyclist = length(casualty_severity[casualty_severity == "Fatal" & casualty_type == "Cyclist"])
            
  )

crash <- left_join(acc, cas_summary, by = "accident_index")
saveRDS(crash,"data/crash_2010_2019_with_summary_adjusted_casualties.Rds")



la_raw <- left_join(la_raw,pop, by = "code")
la_raw <- left_join(la_raw,pop_work_ew, by = "code")

crash <- st_transform(crash, 27700)
la_raw <- st_transform(la_raw, 27700)
```
## Upper Tier Local Authorities and their population

We are using upper tier local authorities for this analysis as they tend to be the local highway authority, thus have overall responsibility for road safety in their areas.

This first plot highlights the significant population and area variation in LAs. 

```{r pressure, echo=FALSE, warning=FALSE, message=FALSE}
m1 <- tm_shape(la_raw) +
  tm_fill(col = "population_2019",
          title = "Population (2019)",
          style = "quantile",
          n = 10, 
          legend.hist = TRUE) +
  tm_borders() +
  tm_layout(legend.outside = TRUE,
            legend.hist.width = 1)

m2 <- tm_shape(la_raw) +
  tm_fill(col = "work_pop_2011",
          title = "Workplace Population (2011)",
          style = "quantile",
          n = 10, 
          legend.hist = TRUE) +
  tm_borders() +
  tm_layout(legend.outside = TRUE,
            legend.hist.width = 1)
tmap_arrange(m1, m2)

```

As LAs vary signifianct is size, population, and character it is unlikely that simple comparing the number of crashes or casualties will be a far comparison. Therefore it would be better to adjust the raw crash rates to account for the local context.

## Crashes by LA in context

```{r crash_la, echo=FALSE, warning=FALSE, message=FALSE}
crash <- st_join(crash, la_raw[,c("la_name", "code")])

# Get Year and Month
crash$year <- year(crash$date)

crash_yr <- crash %>%
  st_drop_geometry() %>%
  group_by(la_name, code, year, ) %>%
  summarise(total_crash = n(),
            crash_fatal = sum(accident_severity == "Fatal"),
            crash_serious = sum(accident_severity == "Serious"),
            crash_slight = sum(accident_severity == "Slight"),
            casualty_slight = sum(casualty_slight),
            casualty_serious = sum(casualty_serious),
            casualty_fatal = sum(casualty_fatal),
            casualty_slight_pedestrian = sum(casualty_slight_pedestrian),
            casualty_serious_pedestrian = sum(casualty_serious_pedestrian),
            casualty_fatal_pedestrian = sum(casualty_fatal_pedestrian),
            casualty_slight_cyclist = sum(casualty_slight_cyclist),
            casualty_serious_cyclist = sum(casualty_serious_cyclist),
            casualty_fatal_cyclist = sum(casualty_fatal_cyclist),
            )

la <- left_join(la_raw, crash_yr[crash_yr$year == 2018,], by = c("la_name", "code"))
la$active_ksi_per100k <- (la$casualty_fatal_pedestrian + la$casualty_fatal_cyclist + la$casualty_serious_pedestrian + la$casualty_serious_cyclist )/ la$population_2019 * 100000
la$active_slight_per100k <- (la$casualty_slight_cyclist +  la$casualty_slight_pedestrian)  / la$population_2019 * 100000

la$active_ksi_per100k_work <- (la$casualty_fatal_pedestrian + la$casualty_fatal_cyclist + la$casualty_serious_pedestrian + la$casualty_serious_cyclist )/ la$work_pop_2011 * 100000

la$active_slight_per_ksi <- (la$casualty_slight_pedestrian + la$casualty_slight_cyclist) /(la$casualty_fatal_pedestrian + la$casualty_fatal_cyclist + la$casualty_serious_pedestrian + la$casualty_serious_cyclist )

la$active_slight_per_ksi[la$active_slight_per_ksi == Inf] <- NA

```

We have adjusted the plots to show the KSI rate of active travelers per 100,000 people (left plot uses resident population; right plot uses workplace population). The KSI rate varies significantly with 90% of LAs being between 8.2 and 24.1 per 100,000 resident population. Scotland has notably low KSI rates, while Cumbria has a high rate.

```{r crash_la_plot, echo=FALSE, warning=FALSE, message=FALSE}
m1 <- tm_shape(la) +
  tm_fill(col = "active_ksi_per100k",
          title = "Active Travel KSI per 100,000 people (2018)",
          style = "quantile",
          n = 10, 
          legend.hist = TRUE) +
  tm_borders() +
  tm_layout(legend.outside = TRUE,
            legend.hist.width = 1)

m2 <- tm_shape(la) +
  tm_fill(col = "active_ksi_per100k_work",
          title = "Active Travel KSI per 100,000 workplace population (2011)",
          style = "quantile",
          n = 10, 
          legend.hist = TRUE) +
  tm_borders() +
  tm_layout(legend.outside = TRUE,
            legend.hist.width = 1)
tmap_arrange(m1, m2)
```

The City of London is an extreme outlier due to its very low resident population. This highlights the denominator issue. The KSI rate is depended on the value chosen as the denominator. In the case of the City of London, the resident population of about 2,500 is not representative of the amount of people traveling by road. A similar affect may be taking place in Cumbria which is a popular tourist destination and may have more active travelers than is represented by its population.

The right plot uses the 2011 census workplace population for England and Wales. The workplace population favors areas like the City of London which have high daytime activity but few nighttime residents. As people working is a subset of all people KSI rates will naturally appear higher. However the relative distribution of risk does change under a workplace population approach suggesting that in at least some areas this is an appropriate way to consider the data.

We can also compare the proportion of slightly injured people vs the proportion of KSI casualties. This we are using the weighting factors to account for the adjusted proportion of serious causalities caused by the historic use of different reporting systems. However this difference may still be due to different reporting practices across the country. 

If this represent a genuine difference in safety, then LAs with a high ratio (dark green) are having proportionally more slight casualties than serious and fatal ones. This could be use to more segregation of active travelers from motor traffic, lower traffic speed or greater driver awareness of active travelers. 


```{r crash_la_plot2, echo=FALSE, warning=FALSE, message=FALSE}
tm_shape(la) +
  tm_fill(col = "active_slight_per_ksi",
          title = "Slight Active casualties per KSI",
          style = "fixed",
          breaks = c(0, 0.5, 1, 1.5,2,2.5,3,3.5,4,10),
          palette = "PRGn",
          midpoint = 2,
          legend.hist = TRUE) +
  tm_borders() +
  tm_layout(legend.outside = TRUE)
```
   

### Top and Bottom LAs

This table lists the top 10 and bottom 10 LAs for Active Travel KSI per 100,000 resident population.

```{r crash_top10, echo=FALSE, warning=FALSE, message=FALSE}
la_top10 <- st_drop_geometry(la)
la_top10 <- la_top10[order(la_top10$active_ksi_per100k, decreasing = TRUE),]
la_top10$rank <- seq(1,nrow(la_top10))
la_top10 <- la_top10[c(1:10,196:205),]
la_top10 <- la_top10[,c("la_name","rank","active_ksi_per100k","total_crash","crash_fatal")]
la_top10$active_ksi_per100k <- round(la_top10$active_ksi_per100k)
knitr::kable(la_top10)
```

As mention above City of London and Westminster are probably distorted due to their high daytime population. This emphasizes the need to adjust for the level of active travel activity not just population.  Ranking by workplace population gives slightly different results.

```{r crash_top10_work, echo=FALSE, warning=FALSE, message=FALSE}
la_top10_work <- st_drop_geometry(la)
la_top10_work <- la_top10_work[order(la_top10_work$active_ksi_per100k_work, decreasing = TRUE),]
la_top10_work$rank <- seq(1,nrow(la_top10_work))
la_top10_work <- la_top10_work[!is.na(la_top10_work$active_ksi_per100k_work),]
la_top10_work <- la_top10_work[c(1:10,seq(nrow(la_top10_work) - 9, nrow(la_top10_work))),]
la_top10_work <- la_top10_work[,c("la_name","rank","active_ksi_per100k_work","total_crash","crash_fatal")]
la_top10_work$active_ksi_per100k_work <- round(la_top10_work$active_ksi_per100k_work)
knitr::kable(la_top10_work)
```

# Historical Trends

Looking at absolute number of KSI, the top 10 and bottom 10 are the same in 2010 and 2019 suggesting no overall change. The plot blow shows the Active Travel KSI rate per 100,000 workplace population from 2010 to 2019. The majority of LAs are shwon with the thin grey lines, while the top 5 and bottom 5 LAs are highlighted in colour.

```{r trends, echo=FALSE, warning=FALSE, message=FALSE}
la <- la[order(la$active_ksi_per100k_work, decreasing = TRUE),]
top_5 <- la$la_name[1:5]
bottom_5 <- la$la_name[seq(169,173)]

crash_yr <- left_join(crash_yr, pop, by = "code")
crash_yr <- left_join(crash_yr, pop_work_ew, by = "code")


crash_yr$active_ksi_per100k <- (crash_yr$casualty_fatal_pedestrian + crash_yr$casualty_fatal_cyclist + crash_yr$casualty_serious_pedestrian + crash_yr$casualty_serious_cyclist )/ crash_yr$population_2019 * 100000

crash_yr$active_ksi_per100k_work <- (crash_yr$casualty_fatal_pedestrian + crash_yr$casualty_fatal_cyclist + crash_yr$casualty_serious_pedestrian + crash_yr$casualty_serious_cyclist )/ crash_yr$work_pop_2011 * 100000

crash_yr$la_name_plot <- ifelse(crash_yr$la_name %in% c(top_5, bottom_5), crash_yr$la_name, "Other" )
crash_yr$lwd <- ifelse(crash_yr$la_name_plot == "Other", FALSE, TRUE )
```

```{r trends_plot, echo=FALSE, warning=FALSE, message=FALSE}

ggplot(crash_yr,
       aes(year, active_ksi_per100k_work, colour = la_name_plot, group = la_name)) +
  geom_line(data = subset(crash_yr, lwd == FALSE), aes(size = lwd)) +
  geom_line(data = subset(crash_yr, lwd == TRUE), aes(size = lwd)) +
  ylab("Active Travel KSI per 100k workplace population") +
  labs(color = "Local Authority") + 
  guides(size = FALSE) +
  guides(colour = guide_legend(override.aes = list(size=3))) +
  scale_size_manual(values = c("TRUE" = 1.5, "FALSE" = 0.1)) +
  scale_color_manual(values=c("#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C", "#FB9A99",
                              "#E31A1C", 
                              "#d3d3d3","#FDBF6F",
                              "#FF7F00", "#CAB2D6", "#6A3D9A")) +
  scale_x_continuous(breaks = 2010:2019, expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  ggtitle("Top and Bottom LAs")

```

The KSI rates fluctuate significantly per year, this reflects the noisy nature of crash data, especially for the subset of crashes that involve an active travelers who is killed or seriously injured. Note that this plot excluded Scotland.

If we look for areas with the biggest change in KSI rates between 2010 and 2019 we can see some weak noisy trends, with areas such a Torbay appreasing to increas in KSIs and Stoke on Trent declining. 

```{r trends2, echo=FALSE, warning=FALSE, message=FALSE}
crash_yr_wide <- pivot_wider(crash_yr[,c("la_name","year","active_ksi_per100k_work")],
                             values_from = "active_ksi_per100k_work",
                             names_from = "year")

crash_yr_wide$change <- (crash_yr_wide$`2019` - crash_yr_wide$`2010`) / crash_yr_wide$`2010` * 100

crash_yr_wide <- crash_yr_wide[order(crash_yr_wide$change, decreasing = TRUE),]

change_top_5 <- crash_yr_wide$la_name[1:5]
change_bottom_5 <- crash_yr_wide$la_name[168:172]

crash_yr$la_name_plot <- ifelse(crash_yr$la_name %in% c(change_top_5, change_bottom_5), crash_yr$la_name, "Other" )
crash_yr$lwd <- ifelse(crash_yr$la_name_plot == "Other", FALSE, TRUE )
```

```{r trends2_plot, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(crash_yr,
       aes(year, active_ksi_per100k_work, colour = la_name_plot, group = la_name)) +
  geom_line(data = subset(crash_yr, lwd == FALSE), aes(size = lwd)) +
  geom_line(data = subset(crash_yr, lwd == TRUE), aes(size = lwd)) +
  ylab("Active Travel KSI per 100k workplace population") +
  labs(color = "Local Authority") + 
  guides(size = FALSE) +
  guides(colour = guide_legend(override.aes = list(size=3))) +
  scale_size_manual(values = c("TRUE" = 1.5, "FALSE" = 0.1)) +
  scale_color_manual(values=c("#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C","#d3d3d3",
                              "#FB9A99", "#E31A1C", "#FDBF6F",
                              "#FF7F00", "#CAB2D6", "#6A3D9A")) +
  scale_x_continuous(breaks = 2010:2019, expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  ggtitle("LAs with Greatest % Change")


```
If we instead compare KSI against the resident population (adding in Scotland to the analysis) we can see that Aberdeen City have made a sustained decline in KSI rates over the last ten years, down over 60% from their high in 2011. Unlike some other LAs where the rate appears variable but steady, the KSI rate in Aberdeen has fallen almost every year and thus may represent a real world change in safety worthy of detailed study. In particular the Launch of Scotland's Road Safety Framework in 2009 may have been successful in reducing Active Travel KSIs.

```{r trends3, echo=FALSE, warning=FALSE, message=FALSE}
crash_yr_wide <- pivot_wider(crash_yr[,c("la_name","year","active_ksi_per100k")],
                             values_from = "active_ksi_per100k",
                             names_from = "year")

# min Max approach din't work
# yr_max <- apply(crash_yr_wide[2:11], 1, FUN=max) 
# yr_min <- apply(crash_yr_wide[2:11], 1, FUN=min)
# crash_yr_wide$change <- (yr_max - yr_min) / yr_min * 100

crash_yr_wide$change <- (crash_yr_wide$`2019` - crash_yr_wide$`2010`) / crash_yr_wide$`2010` * 100
crash_yr_wide <- crash_yr_wide[order(crash_yr_wide$change, decreasing = TRUE),]

change_top_5 <- crash_yr_wide$la_name[1:5]
change_bottom_5 <- crash_yr_wide$la_name[200:204]

crash_yr$la_name_plot <- ifelse(crash_yr$la_name %in% c(change_top_5, change_bottom_5), crash_yr$la_name, "Other" )
crash_yr$lwd <- ifelse(crash_yr$la_name_plot == "Other", FALSE, TRUE )
```

```{r trends3_plot, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(crash_yr,
       aes(year, active_ksi_per100k, colour = la_name_plot, group = la_name)) +
  geom_line(data = subset(crash_yr, lwd == FALSE), aes(size = lwd)) +
  geom_line(data = subset(crash_yr, lwd == TRUE), aes(size = lwd)) +
  ylab("Active Travel KSI per 100k resident population") +
  labs(color = "Local Authority") + 
  guides(size = FALSE) +
  guides(colour = guide_legend(override.aes = list(size=3))) +
  scale_size_manual(values = c("TRUE" = 1.5, "FALSE" = 0.1)) +
  scale_color_manual(values=c("#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C",
                              "#FB9A99", "#E31A1C", "#d3d3d3", "#FDBF6F",  
                              "#FF7F00", "#CAB2D6", "#6A3D9A")) +
  scale_x_continuous(breaks = 2010:2019, expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  ylim(0,30) +
  ggtitle("LAs Change in KSI")
```

## Temporal Analysis

If we consider when active travelers are most likely to be injured, it may provide insights into the causes of these crashes.

```{r tempo, echo=FALSE, warning=FALSE, message=FALSE}
crash$hour <- hour(crash$datetime)
crash$day_of_year <- yday(crash$date)
crash$week_of_year <- week(crash$date)
crash$month_of_year <- month(crash$date)
crash$active_ksi <- crash$casualty_fatal_cyclist + crash$casualty_fatal_pedestrian + crash$casualty_serious_cyclist + crash$casualty_serious_pedestrian
```

```{r tempo_plot, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(crash, aes(hour, active_ksi)) +
  geom_col() +
  ylab("Active Travellers KSI") +
  xlab("Hour of the Day") +
  scale_x_continuous(breaks = 0:24, expand = c(0,0))
```

First we can consider the time of day. This plot show the total number of Active Travelers KSI between 2010 and 2019 for each hour of the day in Great Britain. 
There are clear spikes around the morning and afternoon rush hours and reduced levels during the night. 
This graph largely reflect the general pattern of traffic though the day. 
There are a greater number of active traveller casualties during the PM peak commuting and school run hours (roughly 15:00 - 19:00 PM) as compared to the morning rush hour.
This may perhaps be related to the higher likelihood that people are travelling in darkness or poor light conditions in these hours, or because travellers are more tired at this time.


```{r tempo_plot2, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(crash, aes(day_of_year, active_ksi)) +
  stat_summary(fun.y = sum, geom ='line') +
  ylab("Active Travellers KSI") +
  xlab("Day of the Year") +
  scale_x_continuous(breaks = seq(1,366,30), expand = c(0,0))
```

Consider change through the year, this plot shows the number of Active Travellers KSI between 2010 and 2019 for each day of the year in Great Britain.
The trend is mostly fairly stable with a pronounced low point at Christmas (day 360), when few people travel.
Similarly, rates are relatively low in early April, around the average date of Easter (day 90-100).
KSI rates appear to decline over the summer (days 180 - 240) before rising sharply in September. 
This may reflect a change in travel behavior around the start of the new school year.

The highest active traveller KSI rates are in November and early December (days 300 - 350), which may be related to the early sunsets and dark evenings at this time of year. 
Although the shortest day of the year is typically 21st December, the earliest sunset occurs on 12/13th December.

```{r tempo_plot3, echo=FALSE, warning=FALSE, message=FALSE}
crash_select <- crash[crash$la_name %in% c("City of London","Cumbria","Kingston upon Hull, City of"), ]

ggplot(crash_select, aes(week_of_year, active_ksi, group = la_name, color = la_name)) +
  stat_summary(fun = sum, geom ='line') +
  ylab("Active Travellers KSI") +
  xlab("Week of the Year") +
  scale_x_continuous(breaks = seq(0,53,4), expand = c(0,0))
```

Selecting a few LAs we can see the the distribution of casualties through the year varies a little for place to place. 
Notably, Cumbria's casualties are highest in early August, peak tourism season.

## Hour and day of year

Looking at the distribution of active traveller KSI casualties over both hour and day of year, we can see that the number of casualties in the peak morning hour of 8-9am is high across most of the year, with the exception of the Christmas, Easter and Summer periods.
However, casualties during the hour 5-6pm are highest in Days 300-350 and Days 5-40 (November, December and January, except the Christmas period), suggesting that the darker winter evenings are playing a role here.
Casualties during the hour 4-5pm are highest in Days 320-350, further backing up this observation. 

```{r heatmap}
# there are NAs in the hour. these must be removed first
crash = crash %>% 
  mutate(fiveday = 5*(round(day_of_year/5)))
crash_agg = crash %>%
  st_drop_geometry() %>% 
  select(fiveday, hour, active_ksi) %>% 
  filter(! is.na(hour)) %>% 
  group_by(fiveday, hour) %>% 
  summarise(active_ksi = sum(active_ksi))
summary(crash_agg)
ggplot(crash_agg, aes(x = fiveday, y = hour, fill = active_ksi)) +
  geom_tile() +
  labs(x = "Day of year", y = "Hour")
```

## Hours of darkness

Look at actual light levels
crashes_all$light_conditions

## Looking at commuting times

We can filter out the collisions that occur in non-peak commute hours and look at how the remaining collisions correspond with the LA workplace population.
This better represents the safety of travel to work.
We can see that tourist hotspots like Cumbria now show slightly lower KSI rates, while Lancashire retains a high rate.

```{r commute-hours}
crash_commute_yr <- crash %>%
  st_drop_geometry() %>%
  filter(hour %in% c(7, 8, 9, 16, 17, 18)) %>% 
  group_by(la_name, code, year) %>%
  summarise(total_crash = n(),
            crash_fatal = sum(accident_severity == "Fatal"),
            crash_serious = sum(accident_severity == "Serious"),
            crash_slight = sum(accident_severity == "Slight"),
            casualty_slight = sum(casualty_slight),
            casualty_serious = sum(casualty_serious),
            casualty_fatal = sum(casualty_fatal),
            casualty_slight_pedestrian = sum(casualty_slight_pedestrian),
            casualty_serious_pedestrian = sum(casualty_serious_pedestrian),
            casualty_fatal_pedestrian = sum(casualty_fatal_pedestrian),
            casualty_slight_cyclist = sum(casualty_slight_cyclist),
            casualty_serious_cyclist = sum(casualty_serious_cyclist),
            casualty_fatal_cyclist = sum(casualty_fatal_cyclist),
            )

la_commute <- left_join(la_raw, crash_commute_yr[crash_commute_yr$year == 2018,], by = c("la_name", "code"))
la_commute$active_ksi_per100k <- (la_commute$casualty_fatal_pedestrian + la_commute$casualty_fatal_cyclist + la_commute$casualty_serious_pedestrian + la_commute$casualty_serious_cyclist )/ la_commute$population_2019 * 100000
la_commute$active_slight_per100k <- (la_commute$casualty_slight_cyclist +  la_commute$casualty_slight_pedestrian)  / la_commute$population_2019 * 100000

la_commute$active_ksi_per100k_work <- (la_commute$casualty_fatal_pedestrian + la_commute$casualty_fatal_cyclist + la_commute$casualty_serious_pedestrian + la_commute$casualty_serious_cyclist )/ la_commute$work_pop_2011 * 100000

la_commute$active_slight_per_ksi <- (la_commute$casualty_slight_pedestrian + la_commute$casualty_slight_cyclist) /(la_commute$casualty_fatal_pedestrian + la_commute$casualty_fatal_cyclist + la_commute$casualty_serious_pedestrian + la_commute$casualty_serious_cyclist )

la_commute$active_slight_per_ksi[la_commute$active_slight_per_ksi == Inf] <- NA

la_commute$active_ksi_per100k_work <- (la_commute$casualty_fatal_pedestrian + la_commute$casualty_fatal_cyclist + la_commute$casualty_serious_pedestrian + la_commute$casualty_serious_cyclist )/ la_commute$work_pop_2011 * 100000

m2 <- tm_shape(la_commute) +
  tm_fill(col = "active_ksi_per100k_work",
          title = "Active Travel KSI per 100,000 workplace population (2011)",
          style = "quantile",
          n = 10, 
          legend.hist = TRUE) +
  tm_borders() +
  tm_layout(legend.outside = TRUE,
            legend.hist.width = 1)
m2
```


Commuter hours are classified as 07:00-10:00 and 16:00-19:00.

# Speed limits

Look at roads where the speed limit has changed. Have collision rates changed?

# Focus on Cycling Commuting Times

As we have the best data for cycling activity during commuting times (via the PCT), we have filtered the data in this section to only consider crashes between 7am - 10am and 4pm - 7pm. 

```{r trends_commute, echo=FALSE, warning=FALSE, message=FALSE}
 


```

# Change in cycling

Present the GAM results on change in cycling across England

```{r cycle_trends, echo=FALSE, warning=FALSE, message=FALSE}
 


```

Use the GAM results to look at change in cycling v change in casualties
