---
title: "Local Authority Trend Analysis"
output: github_document
---

THis short analysis is to highlight the variation in crashes affecting active traveling across Great Britain.

Data comes from the Stats19 reports.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sf)
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyr)
library(tmap)
```



```{r get_data, echo = FALSE, include=FALSE}
crash <- readRDS("crashes_active.Rds")
crash <- st_as_sf(crash[!is.na(crash$latitude),], coords = c("longitude","latitude"), crs = 4326)

if(!file.exists("data/counties_uas_gb_2019_ubc.Rds")){
  download.file("https://github.com/saferactive/saferactive/releases/download/0.1.1/counties_uas_gb_2019_ubc.Rds",
              destfile = "data/counties_uas_gb_2019_ubc.Rds", mode = "wb")
}

la <- readRDS("data/counties_uas_gb_2019_ubc.Rds")
la$ctyua19cd[la$ctyua19cd == "E10000002"] <- "E06000060"


pop <- read.csv("data/upper_LA_population_2019.csv")
pop <- pop[pop$mnemonic %in% la$ctyua19cd,]
pop <- pop[,2:3]
names(pop) <- c("code","population_2019")
la <- la[,c("ctyua19cd","ctyua19nm")]
names(la) <- c("code","la_name","geometry")

la <- left_join(la,pop, by = "code")

crash <- st_transform(crash, 27700)
la <- st_transform(la, 27700)
```
## Upper Tier Local Authorities and their population

We are using upper tier local authorities for this analysis as they tend to be the local highway authority, thus have overall responsability for road safety in their areas.

This first plot highlights the significant population and area variation in LAs. 

```{r pressure, echo=FALSE}
tm_shape(la) +
  tm_fill(col = "population_2019",
          title = "Population (2019)",
          style = "quantile",
          n = 10, 
          legend.hist = TRUE) +
  tm_borders() +
  tm_layout(legend.outside = TRUE)
```

## Crashes by LA

```{r crash_la, echo=FALSE, message=FALSE}
crash <- st_join(crash, la[,"la_name"])

# Get Year and Month
crash$year <- year(crash$date)

crash_yr <- crash %>%
  st_drop_geometry() %>%
  group_by(la_name, year) %>%
  summarise(total_crash = n(),
            crash_fatal = sum(accident_severity == "Fatal"),
            crash_serious = sum(accident_severity == "Serious"),
            crash_slight = sum(accident_severity == "Slight"),
            cas_fatal = sum(number_of_casualties[accident_severity == "Fatal"]),
            cas_serious = sum(number_of_casualties[accident_severity == "Serious"]),
            cas_slight = sum(number_of_casualties[accident_severity == "Slight"])
            )

la <- left_join(la, crash_yr[crash_yr$year == 2018,], by = c("la_name"))
la$ksi_per100k <- la$cas_fatal + la$cas_serious / la$population_2019 * 100000
la$slight_per100k <- la$cas_slight  / la$population_2019 * 100000
```

We can adjust the plots the KSI rate for population, England is clearly doing worse than Scotland and Wales, particularly in the home counties. No this is a count based on the most serious causality, so may slightly overestimate the KSI rate. The City of London is an extreme outlier due to it very low resident population.

```{r crash_la_plot, echo=FALSE}
tm_shape(la) +
  tm_fill(col = "ksi_per100k",
          title = "KSI per 100,000 people (2018)",
          style = "quantile",
          n = 10,
          legend.hist = TRUE) +
  tm_borders() +
  tm_layout(legend.outside = TRUE)
```
The pattern of slight casualties is different to the KSI

```{r crash_la_plot2, echo=FALSE}
tm_shape(la) +
  tm_fill(col = "slight_per100k",
          title = "Slight Casualties per 100,000 people (2018)",
          style = "quantile",
          n = 10,
          legend.hist = TRUE) +
  tm_borders() +
  tm_layout(legend.outside = TRUE)
```


```{r crash_top10, echo=FALSE}
la_top10 <- st_drop_geometry(la)
la_top10 <- la_top10[order(la_top10$ksi_per100k, decreasing = TRUE),]
la_top10$rank <- seq(1,nrow(la_top10))
la_top10 <- la_top10[c(1:10,196:205),]
la_top10 <- la_top10[,c("la_name","rank","ksi_per100k","total_crash","crash_fatal")]
la_top10$ksi_per100k <- round(la_top10$ksi_per100k)
la_top10
```

City of London and Westminster are probably distorted due to there high daytime population. This emphasizes the need to adjust for the level of active travel activity not just population. 

# Historical Trends

Looking at absolute number of KSI, the top 10 and bottom 10 are the same in 2009 and 2018 suggesting no overall change.

```{r trends, echo=FALSE}
rank_2018 <- crash_yr[crash_yr$year == 2018,]
rank_2018 <- rank_2018[order(rank_2018$total_crash, decreasing = TRUE),]
top_10_2018 <- rank_2018$la_name[1:10]
bottom_10_2018 <- rank_2018$la_name[seq(nrow(rank_2018) - 9,nrow(rank_2018))]

rank_2009 <- crash_yr[crash_yr$year == 2018,]
rank_2009 <- rank_2009[order(rank_2009$total_crash, decreasing = TRUE),]
top_10_2009 <- rank_2009$la_name[1:10]
bottom_10_2009 <- rank_2009$la_name[seq(nrow(rank_2009) - 9,nrow(rank_2009))]

ggplot(crash_yr[crash_yr$la_name %in% top_10_2018,],
       aes(year, total_crash, colour = la_name)) +
  geom_line() +
  ylab("Crashes with active travel casualties") +
  labs(color = "Local Authority") +
  ggtitle("Top 10 LAs in 2018")

ggplot(crash_yr[crash_yr$la_name %in% bottom_10_2018,],
       aes(year, total_crash, colour = la_name)) +
  geom_line() +
  ylab("Crashes with active travel casualties") +
  labs(color = "Local Authority") +
  ggtitle("Bottom 10 LAs in 2018") 


```

If we look for areas with the biggest change

```{r trends2, echo=FALSE}
crash_yr_wide <- pivot_wider(crash_yr[,c("la_name","year","total_crash")],
                             values_from = "total_crash",
                             names_from = "year")

crash_yr_wide$change <- (crash_yr_wide$`2018` - crash_yr_wide$`2009`) / crash_yr_wide$`2009` * 100

crash_yr_wide <- crash_yr_wide[order(crash_yr_wide$change, decreasing = TRUE),]

change_top_10 <- crash_yr_wide$la_name[1:10]
change_bottom_10 <- crash_yr_wide$la_name[seq(nrow(crash_yr_wide) - 9,nrow(crash_yr_wide))]


ggplot(crash_yr[crash_yr$la_name %in% change_top_10,],
       aes(year, total_crash, colour = la_name)) +
  geom_line() +
  ylab("Crashes with active travel casualties") +
  labs(color = "Local Authority") +
  ggtitle("10 LAs with greatest % increase in crashes")

ggplot(crash_yr[crash_yr$la_name %in% change_bottom_10,],
       aes(year, total_crash, colour = la_name)) +
  geom_line() +
  ylab("Crashes with active travel casualties") +
  labs(color = "Local Authority") +
  ggtitle("10 LAs with greatest % decrease in crashes")


```
There is a clear lesson that Scotland has done something significant in terms of reducing KSI of active travelers especially in Aberdeen. At this stage we cannot show if this is a true change in the crash rate or a change in recording practices. Due to the gradual an continuous decline it seems unlikely to be caused by a chang in data gatherin policy. It may have been caused by a reduction in active travel.


