---
title: "SaferActive"
subtitle: "📉🚂🚅🚎🛴🚲🚶📈<br/>(Visual) analysis of road safety data"
author: "The SaferActive Team"
institute: "University of Leeds, [Institute for Transport Studies](https://environment.leeds.ac.uk/transport)"
date: "University of Leeds, 2021-10-15 (updated: `r Sys.Date()`)\nReproducible source code: [github.com/saferactive/saferactive](https://github.com/saferactive/saferactive)"
output:
  xaringan::moon_reader:
    css: [default, robot, robot-fonts]
    lib_dir: libs
    nature:
      ratio: '16:9'
      beforeInit: "https://platform.twitter.com/widgets.js"
      highlightStyle: github
      highlightLines: false
      countIncrementalSlides: false
---

```{r, include=FALSE}
library(tidyverse)
library(kableExtra)
```

### Navigating the Stats19 data analysis space

???

starting point : hard to generate genuine/solid findings

large-scale observational dataset -- analysis space large.
So difficult to know where to start.

*Are* examples of CDA aiming to establish specific things from known data.

Here -- large set of variables; their structure not always known about; sometimes difficult to establish expectations.

Data analysis to make decisions!

--

.pull-left[
.right[**`accidents`**]
.right[`accident_index ..`]
]
.pull-right[
crash time, location (~10m resolution), admin location, road context and weather and light conditions
]

.pull-left[
.right[**`casualties`**]
.right[`accident_index  ..`]
.right[`vehicle_reference  ..`]
.right[`casualty_reference  ..`]
]
.pull-right[
casualty sex, age, severity, passenger type, home area demogs
]

.pull-left[
.right[**`vehicles`**]
.right[`accident_index  ..`]
.right[`vehicle_reference  ..`]
]
.pull-right[
vehicles type, age, capacity, manoeuvre type, driver sex, age,  home area demogs, purpose
]


???

So this is where exploratory analysis and vis approaches are suited -- but EDA and vis not all about *discovery* -- describing data, exploring patterns and deriving findings.

---

### Science reform and retiring the EDA/CDA distinction

<img src = "img/challenges_eda.png", width = 55%, style = "position:absolute; top: 20%; left: 8%;"></img>

<img src = "img/harvard_challenges_eda.png", width = 20%, style = "position:absolute; top: 20%; left: 70%;"></img>

???

That about visual data analysis as model formulation and critique : about helping to form comparisons and expectations -- and characterising variation from those expectations.

Also about helping to guard against false discovery -- or at least that should be our ambition.

---


### Framework for formalising analysis


<img src = "img/framework.png", width = 50%, style = "position:absolute; top: 18%; left: 10%;"></img>


???

Net of the dominant pattern

Data analysis for decision-making

---


### Pedestrian-Vehicle crashes : who-hit-whom demographics

???
Research identifying diverging characteristics of drivers and those being hit by them.

--


.pull-left[
.right[**`accidents`**]
.right[`accident_index ..`]
]
.pull-right[
crash time, **location (~10m resolution)**, admin location, road context and weather and light conditions
]

.pull-left[
.right[**`casualties`**]
.right[`accident_index  ..`]
.right[`vehicle_reference  ..`]
.right[`casualty_reference  ..`]
]
.pull-right[
casualty sex, age, severity, passenger type, **home area demogs**
]

.pull-left[
.right[**`vehicles`**]
.right[`accident_index  ..`]
.right[`vehicle_reference  ..`]
]
.pull-right[
vehicles type, age, capacity, manoeuvre type, driver sex, age,  **home area demogs**, purpose
]

---

### Pedestrian-Vehicle crashes : who-hit-whom demographics

Crash characteristics:

* **`accidents`** : IMD class of crash location : {1 high; 5 low}
* **`casualties`** : IMD class of pedestrian : {1 high; 5 low}
* **`vehicles`** : IMD class of driver : {1 high; 5 low}


--

<img src = "img/imd-driver-cas-title.png", width = 50%, style = "position:absolute; top: 38%; left: 10%;"></img>

<img src = "img/imd-driver-cas-obs.png", width = 15%, style = "position:absolute; top:46%; left: 10%;"></img>

???

---

### Pedestrian-Vehicle crashes : who-hit-whom demographics

Crash characteristics:

* **`accidents`** : IMD class of crash location : {1 high; 5 low}
* **`casualties`** : IMD class of pedestrian : {1 high; 5 low}
* **`vehicles`** : IMD class of driver : {1 high; 5 low}



<img src = "img/imd-driver-cas-title.png", width = 50%, style = "position:absolute; top: 38%; left: 10%;"></img>

<img src = "img/imd-driver-cas-obs.png", width = 13%, style = "position:absolute; top:46%; left: 10%;"></img>

--

<img src = "img/imd-driver-cas-full.png", width = 50%, style = "position:absolute; top: 70%; left: 10%;"></img>

--

<img src = "img/imd-driver-cas-resid.png", width = 16.5%, style = "position:absolute; top: 46%; left: 10%;"></img>

???

We expect independence in the relative number recorded crashes by IMD group of pedestrian and driver -- so given the row_totals -- the frequencies by each IMD of driver -- and col_totals -- the frequencies by each IMD  of pedestrian -- we spread expectation amongst the contingency table assuming that joint driver-cas frequencies by IMD are independent.

When we do that -- try to control for dominant pattern -- we find that heavy association between demogs: those more demographically similar are more likely to crash with each other. Particularly true -- darker colour -- for low deprivation driver-cas and might suggest that sphere of circulation/mixing is more demographically narrow. Certainly cities tend not to contain so often very low deprivation.

So controlling for dominant pattern helps us identify obvious confounder.



---

### Pedestrian-Vehicle crashes : who-hit-whom demographics

Crash characteristics:

* **`accidents`** : IMD class of **crash location** : {1 high; 5 low}
* **`casualties`** : IMD class of pedestrian : {1 high; 5 low}
* **`vehicles`** : IMD class of driver : {1 high; 5 low}

--

<img src = "img/imd-driver-cas-crash-title.png", width = 55%, style = "position:absolute; top: 38%; left: 10%;"></img>

<img src = "img/imd-driver-cas-crash-obs.png", width = 55%, style = "position:absolute; top: 46%; left: 10%;"></img>


???

Splitting by crash location exposes this

Spread expectation across the contingency table in the same way -- independent of driver-casualty demographics.
---


### Pedestrian-Vehicle crashes : who-hit-whom demographics

Crash characteristics:

* **`accidents`** : IMD class of **crash location** : {1 high; 5 low}
* **`casualties`** : IMD class of pedestrian : {1 high; 5 low}
* **`vehicles`** : IMD class of driver : {1 high; 5 low}


<img src = "img/imd-driver-cas-crash-title.png", width = 55%, style = "position:absolute; top: 38%; left: 10%;"></img>

<img src = "img/imd-driver-cas-crash-resid.png", width = 55%, style = "position:absolute; top: 46%; left: 10%;"></img>



---

### Pedestrian-Vehicle crashes : who-hit-whom demographics

Crash characteristics:

* **`accidents`** : IMD class of **crash location** : {1 high; 5 low}
* **`casualties`** : IMD class of **pedestrian** : {1 high; 5 low}
* **`vehicles`** : IMD class of **driver** : {1 high; 5 low}

<img src = "img/imd-geodemog-dist-title.png", width = 55%, style = "position:absolute; top: 38%; left: 10%;"></img>

<img src = "img/imd-geodemog-dist.png", width = 55%, style = "position:absolute; top: 46%; left: 10%;"></img>


---

### Pedestrian-Vehicle crashes : who-hit-whom demographics

Crash characteristics:

* **`accidents`** : IMD class of **crash location** : {1 high; 5 low}
* **`casualties`** : IMD class of **pedestrian** : {1 high; 5 low}
* **`vehicles`** : IMD class of **driver** : {1 high; 5 low}

<img src = "img/imd-geodemog-dist-title.png", width = 55%, style = "position:absolute; top: 38%; left: 10%;"></img>


<img src = "img/imd-driver-cas-crash-dist-full.png", width = 55%, style = "position:absolute; top: 46%; left: 10%;"></img>

---


### Pedestrian-Vehicle crashes : who-hit-whom demographics

Crash characteristics:

* **`accidents`** : IMD class of **crash location** : {1 high; 5 low}
* **`casualties`** : IMD class of **pedestrian** : {1 high; 5 low}
* **`vehicles`** : IMD class of **driver** : {1 high; 5 low}

<img src = "img/imd-geodemog-dist-title.png", width = 55%, style = "position:absolute; top: 38%; left: 10%;"></img>

<img src = "img/imd-driver-cas-crash-dist-resid.png", width = 55%, style = "position:absolute; top: 46%; left: 10%;"></img>


---

### Pedestrian-Vehicle crashes : who-hit-whom severity

--

<img src = "img/bars-assoc-title.png", width = 25%, style = "position:absolute; top: 25%; left: 10%;"></img>

<img src = "img/bar-freq.png", width = 25%, style = "position:absolute; top: 33%; left: 10%;"></img>

---

### Pedestrian-Vehicle crashes : who-hit-whom severity

<img src = "img/bars-assoc-title.png", width = 25%, style = "position:absolute; top: 25%; left: 10%;"></img>

<img src = "img/bars-assoc-stacked.png", width = 25%, style = "position:absolute; top: 32%; left: 10%;"></img>

--

<img src = "img/bars-assoc-exp.png", width = 22.2%, style = "position:absolute; top: 32%; left: 35.5%;"></img>

--

<img src = "img/bars-assoc-prop.png", width = 22.3%, style = "position:absolute; top: 32%; left: 35.5%;"></img>

--

<img src = "img/bars-assoc-mosaic.png", width = 26%, style = "position:absolute; top: 32%; left: 58%;"></img>

---

### Pedestrian-Vehicle crashes : who-hit-whom severity

**Expectation**

*---- Casualty severity does not vary depending on vehicle type*

--
```{r tab-vehicle-severity-resids, echo=FALSE, message=FALSE, out.width="90%"}
readr::read_csv("img/vehicle_severity_cross_resids.csv") %>%
  kbl(caption = "") %>%
  row_spec(9, bold=T) %>% column_spec(4,bold = T) %>%
  add_header_above(c(" ", "Observed" = 3, "Expected" = 2, "Signed chi-scores"=2))
```
???

Chi-square residuals


---

### Pedestrian-Vehicle crashes : who-hit-whom severity


**Expectation**

*---- Casualty severity **and crash frequency** does not vary depending on vehicle type **and location***

--

<img src = "img/mosaic_harrow_westminster.png", width = 55%, style = "position:absolute; top: 35%; left: 10%;"></img>


--

<img src = "img/mosaic_harrow_westminster_colour.png", width = 55%, style = "position:absolute; top: 35%; left: 10%;"></img>

--

<img src = "img/mosaic_harrow_westminster_colour_none.png", width = 55%, style = "position:absolute; top: 35%; left: 10%;"></img>

---

### Pedestrian-Vehicle crashes : who-hit-whom severity

**Expectation**

*---- Casualty severity **and crash frequency** does not vary depending on vehicle type **and location***


<img src = "img/anim_real_grid.gif", width = 45%, style = "position:absolute; top: 35%; left: 8%;"></img>


---

### Pedestrian-Vehicle crashes : who-hit-whom severity

**Expectation**

*---- Casualty severity **and crash frequency** does not vary depending on vehicle type **and location***

<img src = "img/mosaic_boroughs_spatial.png", width = 42%, style = "position:absolute; top: 35%; left: 8%;"></img>

---

### Pedestrian-Vehicle crashes : who-hit-whom severity

**Expectation**

*---- Casualty severity **and crash frequency** does not vary depending on vehicle type **and location***

<img src = "img/mosaic_boroughs_spatial_simp.png", width = 42%, style = "position:absolute; top: 35%; left: 8%;"></img>

---

### Pedestrian-Vehicle crashes : who-hit-whom severity

**Expectation**

*---- Casualty severity does not vary depending on location*

<img src = "img/bars_boroughs_severity.png", width = 42%, style = "position:absolute; top: 35%; left: 8%;"></img>

---

### Pedestrian-Vehicle crashes : who-hit-whom severity

**Expectation**

*---- Casualty severity does not vary depending on location*

<img src = "img/boroughs_car.png", width = 42%, style = "position:absolute; top: 35%; left: 8%;"></img>

---

### Pedestrian-Vehicle crashes : who-hit-whom severity

**Expectation**

*---- Casualty severity does not vary depending on location*

<img src = "img/lineup.png", width = 55%, style = "position:absolute; top: 35%; left: 8%;"></img>

---


### Pedestrian-Vehicle crashes : **monitoring** severity by area

--

#### Where, and by how much, are injury severity rates high or low?

--
.pull-left[
.right[risk-ratio<sub>lad</sub> = ksi<sub>lad</sub> / ksi<sub>nat</sub>]
]
--

<img src = "img/rrs-title.png", width = 35%, style = "position:absolute; top: 14%; left: 60%;"></img>

<img src = "img/rrs.png", width = 35%, style = "position:absolute; top: 20%; left: 60%;"></img>

--
<br><br><br>
.pull-left[
 .right[High : Uttlesford 83%]  <br>
 .right[Low : Bromsgrove 4%]
]
???
---
### Pedestrian-Vehicle crashes : **monitoring** severity by area


#### Where, and by how much, are injury severity rates high or low?

.pull-left[
.right[risk-ratio<sub>lad</sub> = ksi<sub>lad</sub> / ksi<sub>nat</sub>]
]


<img src = "img/rrs-title.png", width = 35%, style = "position:absolute; top: 14%; left: 60%;"></img>

<img src = "img/rrs.png", width = 35%, style = "position:absolute; top: 20%; left: 60%;"></img>

<br><br><br>
.pull-left[
 .right[High : Uttlesford 83% -- **12 crashes**]  <br>
 .right[Low : Bromsgrove 4% -- **27 crashes**]
]


???

---

### Pedestrian-Vehicle crashes : **monitoring** severity by area


#### Where, and by how much, are injury severity rates high or low?


.pull-left[
.right[risk-ratio<sub>lad</sub> = ksi<sub>lad</sub> / ksi<sub>nat</sub>]
]


<img src = "img/rrs-title-boot.png", width = 35%, style = "position:absolute; top: 14%; left: 60%;"></img>

<img src = "img/rrs-boot.png", width = 35%, style = "position:absolute; top: 20%; left: 60%;"></img>

<br><br><br>
.pull-left[
 .right[High : Uttlesford 83% -- **12 crashes**]  <br>
 .right[Low : Bromsgrove 4% -- **27 crashes**]
]

--


<img src = "img/empirical-bayes.png", width = 45%, style = "position:absolute; top: 40%; left: 30%;"></img>

---
### Pedestrian-Vehicle crashes : **monitoring** severity by area


#### Where, and by how much, are injury severity rates high or low?

.pull-left[
.right[risk-ratio<sub>lad</sub> = ksi<sub>lad</sub> / ksi<sub>nat</sub>]
]


<img src = "img/rrs-title-boot.png", width = 35%, style = "position:absolute; top: 14%; left: 60%;"></img>

<img src = "img/rrs-boot.png", width = 35%, style = "position:absolute; top: 20%; left: 60%;"></img>


<br><br><br>
.pull-left[
 .right[High : Uttlesford 83% -- 12 crashes **> 58%**]  <br>
 .right[Low : Bromsgrove 4% -- 27 crashes **> 13%**]
]



???

---

### Pedestrian-Vehicle crashes : **monitoring** severity by area


#### Where, and by how much, are injury severity rates high or low?

.pull-left[
.right[risk-ratio<sub>lad</sub> = ksi<sub>lad</sub> / ksi<sub>nat</sub>]
]


<img src = "img/rrs-title-boot.png", width = 35%, style = "position:absolute; top: 14%; left: 60%;"></img>

<img src = "img/rrs-boot.png", width = 35%, style = "position:absolute; top: 20%; left: 60%;"></img>


<br><br><br>
.pull-left[
.right[High : Uttlesford 83% -- 12 crashes **> 58%**]  <br>
.right[Low : Bromsgrove 4% -- 27 crashes **> 13%**]  <br> <br>
 .right[**Multiple testing**?]
]



---


### Pedestrian-Vehicle crashes : **monitoring** severity by area



#### So how risky is it when a pedestrian is hit by a vehicle?

--
.pull-left[
 .right[Bristol 15%, 228 crashes]
]

.pull-right[
.left[Sheffield 50%, 248 crashes]
]

--

<img src = "img/risk-theatre.jpg", width = 50%, style = "position:absolute; top: 45%; left: 25%;"></img>

???

2% | 15% | 28%

It is notoriously difficult to develop intuition around these sorts of relative frequencies. In the STATS19 dataset, we might wish to compare the injury severity rate of pedestrain-vehicle road crashes – the proportion of all crashes that resulted in a serious injury or fatality (KSI) – taking place between two local authority areas, say Bristol and Sheffield. There is in fact quite a difference in the injury severity rate between these two local authority areas in 2019: 15% for Bristol (35 out of 228 reported crashes were KSI) versus 50% for Sheffield (124 out of 248 reported crashes were KSI).

---


### Pedestrian-Vehicle crashes : **monitoring** severity by area



#### Where, and by how much, are injury severity rates high or low?

#### So how risky is it when a pedestrian is hit by a vehicle?

.pull-left[
 .right[Bristol 15%, 228 crashes]
]

.pull-right[
.left[Sheffield 50%, 248 crashes]
]


<img src = "img/icon-arrays.png", width = 45%, style = "position:absolute; top: 45%; left: 27%;"></img>

---

### Framework for formalising analysis


<img src = "img/framework.png", width = 50%, style = "position:absolute; top: 18%; left: 10%;"></img>


???

Net of the dominant pattern

Data analysis for decisions
